---
title: "Snapshot Conversion: From JSON to Parquet"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Snapshot Conversion: From JSON to Parquet}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  cache: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

`openalexSnapshot` provides three functions for processing the OpenAlex bulk snapshot:

| Task | Function |
|------|----------|
| Convert JSON to Parquet | `snapshot_to_parquet()` |
| Build ID lookup index | `build_corpus_index()` |
| Extract records by ID | `lookup_by_id()` |

All three functions are backed by a compiled Rust library (via extendr) that is statically linked into the package. No external binary is required. Precompiled binaries for macOS and Linux are available on r-universe:

```{r}
install.packages(
  "openalexSnapshot",
  repos = c("https://rkrug.r-universe.dev", "https://cloud.r-project.org")
)
```

## Root Directory Layout

All three functions use a single `root_dir` parameter. All paths are derived from it:

```
<root_dir>/
  snapshot/                          # raw JSON .gz files (input)
  parquet/                           # converted Parquet files (output)
  parquet/<dataset>_id_idx.parquet   # lookup index
  openalex-snapshot_metadata/        # metadata and schema caches
```

# `snapshot_to_parquet()` -- Convert JSON to Parquet

```{r}
library(openalexSnapshot)

# Convert all datasets (works, authors, institutions, ...)
snapshot_to_parquet(
  root_dir     = "/Volumes/openalex",
  workers      = 4,
  profile      = "balanced",   # "safe" | "balanced" | "fast"
  memory_limit = 15000         # MB
)

# Convert a specific dataset only
snapshot_to_parquet(
  root_dir  = "/Volumes/openalex",
  data_sets = "works",
  workers   = 4
)
```

`data_sets = NULL` (default) processes all datasets. When a character vector is given, only the specified datasets are converted.

The function returns `invisible(root_dir)`.

## Performance Profiles

The `profile` argument controls the speed/memory trade-off:

| Profile | Memory use | Speed | When to use |
|---------|-----------|-------|-------------|
| `"safe"` | Low | Slow | Shared machines, limited RAM |
| `"balanced"` | Medium | Medium | Default for most workloads |
| `"fast"` | High | Fast | Dedicated machines with >= 32 GB RAM |

## Resume Support

`snapshot_to_parquet()` skips already-converted files, so it is safe to re-run after an interruption.

# `build_corpus_index()` -- Build ID Lookup Indexes

After conversion, build an ID index for fast record retrieval:

```{r}
build_corpus_index(
  root_dir  = "/Volumes/openalex",
  data_sets = NULL,          # NULL = all datasets
  workers   = 8,
  profile   = "balanced",
  overwrite = FALSE          # TRUE = rebuild even if index exists
)
```

This creates `<root_dir>/parquet/<dataset>_id_idx.parquet` for each dataset.

## How the Index Works

The index contains four columns:

| Column | Description |
|--------|-------------|
| `id` | Full OpenAlex URL (e.g. `https://openalex.org/W2741809807`) |
| `id_block` | `floor(numeric_id / 10000)` -- used for block-based partitioning |
| `parquet_file` | Relative path to the source Parquet file |
| `file_row_number` | 0-indexed row within that file |

The `id_block` grouping allows the lookup function to filter the index to just the relevant partition before reading source files, avoiding a full index scan.

## Two-Stage Index Build

**Stage 1**: Each Parquet file is indexed individually (one connection per worker), writing a temporary per-file result to a staging directory.

**Stage 2**: The staging directory is merged into a single `*_id_idx.parquet` file. The staging directory is then removed.

# `lookup_by_id()` -- Extract Records by OpenAlex ID

```{r}
# Extract specific works
out_dir <- tempfile()
lookup_by_id(
  index_file = "/Volumes/openalex/parquet/works_id_idx.parquet",
  ids        = c("W2741809807", "W2100837269", "W1234567890"),
  output_dir = out_dir
)

# Read the results with arrow
library(arrow)
works <- open_dataset(out_dir) |> collect()
```

IDs can be in short form (`"W2741809807"`) or long form (`"https://openalex.org/W2741809807"`). Both are accepted and normalised automatically.

## How Lookup Works

1. IDs are normalised to long form (`https://openalex.org/...`).
2. The index file is filtered to find the source file and row number for each ID.
3. Results are grouped by source Parquet file.
4. For each source file, only the required rows are retrieved -- no full-file scan.
5. Results are written to `output_dir` as Parquet files.

## Using the Results

The output directory contains Parquet files that can be read with `arrow`:

```{r}
library(arrow)

works <- open_dataset(out_dir) |> collect()
nrow(works)
```

For further analysis with DuckDB:

```{r}
library(duckdb)
library(DBI)

con <- dbConnect(duckdb())
dbExecute(con, paste0(
  "CREATE VIEW extracted AS SELECT * FROM parquet_scan('",
  out_dir, "/*.parquet')"
))
result <- dbGetQuery(con, "SELECT id, title, publication_year FROM extracted")
dbDisconnect(con)
```

# Worked Example

A minimal end-to-end pipeline:

```{r}
library(openalexSnapshot)
library(arrow)

root <- "/Volumes/openalex"

# Step 1: Convert the snapshot
snapshot_to_parquet(root_dir = root, workers = 4, memory_limit = 15000)

# Step 2: Build the index for works
build_corpus_index(root_dir = root, data_sets = "works", workers = 4)

# Step 3: Look up specific records
out_dir <- file.path(root, "my_extract")
lookup_by_id(
  index_file = file.path(root, "parquet", "works_id_idx.parquet"),
  ids        = c("W2741809807", "W2100837269"),
  output_dir = out_dir
)

# Step 4: Read results
works <- open_dataset(out_dir) |> collect()
print(works[, c("id", "title", "publication_year")])
```
